The Essence of Summarization\\ in Pushdown Flow Analyses\\ \textbf{Supplemental Material}

### 2014-03-20 content.tex
Bugfix in system definition. Initial state was added to R' (type error) instead of F'.
Filling out correctness proof in metatheory. Still needs a bit more work (and discussion of order invariants)

%%
Testing new ideas in analysis for improving precision or increasing performance can be a difficult venture.
%
Premature abstraction can mask correctness issues with complicated machinery we employ in our analyses\footnote{Oh so many personal experiences}.
%
We contend that analysis machinery should have an executable concrete counterpart that maintains the meaning of the language for effective testing.
%
Abstraction should then be a simple process that is ``obviously correct.''
%
The framework we present in this paper is applicable in the concrete such that a point-wise abstraction leads to the pushdown analyses in the literature.
%%

@@

Constructing models and analyzing models are different processes.
%
The goal of this work is to simplify the \textit{program} $\to$ \textit{model} process.
%
First-order languages have the luxury of a fully-precise control flow graph that can be extracted from a program AST in linear time.
%
In a sense, the first-order world gets their models ``for free.''
%
Higher-order recursion schemes (HORS) are a powerful model for simply typed functional programs with good model-checking properties~\citep{dvanhorn:Ong2006ModelChecking}, but the \textit{program} $\to$ \textit{HORS} process is narrowly applicable.
%
It is still a matter of mystery for HORS to support programming language features like mutable state and untyped computation.
%%
As a middle ground, we show a simple way to transform one's programming language semantics into a what can be viewed as a pushdown model constructor.
%%

@@

CFA2 is an online version of a technique called ``summarization'' from Sharir and Pnueli \citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with PDCFA's crucial ``$\epsilon$-closure graph''.
%
The online quality of the algorithm is what makes it suitable for higher-order languages, since CFA essentially interprets the program for each discovered state to build the model on the fly.
%
Summarization algorithms need not be restricted to languages with well-bracketed calls and returns.
%
The technique is adaptable for higher precision in the common case but to still handle difficult cases such as first-class control.
%
This was shown for the \rackett{call/cc} operator in~\citet{ianjohnson:Vardoulakis2011Pushdown}.
%
This work illuminated the fact that we can harness the enhanced technology of pushdown analyses in non-pushdown models of computation.
%
Doing this sacrifices call/return matching in the general case, but in practice the precision is much better than the alternative regular model like 0-CFA would provide.
%%

@@

%%
The second piece of machinery serves two purposes:
%
(1) the state space contains all function callsites to project out a control-flow graph, and %
(2) the worklist semantics of execution makes it possible to appeal to an incomplete memo table, so the saved stacks allow new results to flow to calls that used incomplete memo information.
%
Consider a non-deterministic semantics where a function call reaches a result and memoizes, a second call uses the memo table (storing its stack), and then the original call reaches another result (since the worklist allows for such parallel reduction).
%
The second result had better also flow to the second call to maintain soundness!
%%

%%
Once the machine semantics is in this form, simple point-wise abstraction leads to the summarization algorithms that we see in the literature.
%
In \autoref{sec:pdcfa} we derive a cousin of PDCFA.
%
The following section shows how to extend PDCFA with abstract garbage collection. % \autoref{sec:gc}: we show how abstract garbage collection ``falls out'' of this approach.
%
Then in \autoref{sec:cfa2} we show the fundamental difference between PDCFA and CFA2 without first-class control. % : we make additions to the PDCFA semantics to get a direct-style CFA2 without first-class control}
%
\ifwcm{We show in \autoref{sec:wcm} that semantics with stack-inspection (beyond GC) also fit within this interpretation.}
%
Finally, we show in \autoref{sec:sr} that even though first-class control operators don't fit in the pushdown model, we can gracefully loosen the abstraction to analyze delimited, composable first-class control.

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}%{r|ll}
    \hline
% variable lookup
    $\tpl{\svar\mvar, \menv, \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
%    & \textsc{[variable lookup]}
    \\
% application
    $\tpl{\sapp{\mexpri0}{\mexpri1}, \menv, \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mexpri0, \menv, \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}, \mktab, \mmemo}$
%    & \textsc{[function eval]}
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{\mexpr, \menv, \mstore, \kcons{\kfn{\mval}}{\mkont}, \mktab, \mmemo}$
%    & \textsc{[argument eval]}
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\slam{\mvar}{\mexpr},\menv}}{\mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \mktab',
          \mmemo}$
%    & \textsc{[un-memoized call]}
     \\ & \quad if $(\mpoint, \mstore') \notin \dom(\mmemo)$
\\
    & or \\
    & $\tpl{\mval_\mathit{result},
            \mstore'',
            \mkont,
            \mktab',
            \mmemo}$ % & \textsc{[memo lookup]}
    \\ & \quad if $(\mval_\mathit{result},\mstore'') \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
          & $\mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{\mkont}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, M'}$
%    & \textsc{[memoize/return]}
    \\ & \quad if $\mkont \in \mktab(\mpoint, \mstore')$
    \\ where & $\mmemo' = \joinone{\mmemo}{(\mpoint, \mstore')}{(\mval,\mstore)}$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

This semantics is only correct for fresh allocation strategies, which in turn make the reduction relation deterministic.
%
As it is, the CESK machine simulates the tabular machine, but not the other way around;
%
if we use a finite allocation strategy, then we don't explore some non-deterministic paths that lead to different results for a function call.
%
We thus explore all paths and share $\mktab$ and $\mmemo$ between all paths.

@@

{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mastate \in \sa{State} &::= \tpl{\mexpr,\menv,\mstore,\mkont} \\
  \System &= \wp(\sa{State}) \times \wp(\sa{State} \times \sa{State}) \times \KTab \times \Memo \\
  {\mathcal F} &: \sa{\State} \to \System \to \System
\end{align*}
\begin{align*}
  {\mathcal F}(\mastate_0)(\emptyset, R, \mktab, \mmemo) &= (\emptyset, R, \mktab, \mmemo) \\
  {\mathcal F}(\mastate_0)(F\uplus\set{\mastate}, R, \mktab, \mmemo) &= (F\cup F', R \cup R', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\wn(\mastate,\mktab,\mmemo) \stepto \mstate'} \\
                \wn(\tpl{\mexpr,\menv,\mstore,\mkont},\mktab,\mmemo)
                   &= \tpl{\mexpr,\menv,\mstore,\mkont,\mktab,\mmemo}
\end{align*}
\begin{align*}
    R' &= \setbuild{(\mastate,\mastate')}{\wn(\mastate', \_, \_) \in I \wedge \mastate\nequiv\tpl{\mval,\mstore,\krt{\mctx}}} \cup \set{\mastate_0}\\
    \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \mktab', \_) \in I} \\
    \mmemo' &= \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \mmemo') \in I} \\
    F' &= \setbuild{\mastate'}{\wn(\mastate', \_, \_) \in I} \setminus (\pi_1 R \cup \pi_2 R)
  \end{align*}
}
%%

@@

Garbage collection ($\Gamma$) restricts the heap to just the reachable addresses, so a semantics with GC introduces the following reduction rule:
\begin{center}
  \begin{minipage}{0.3\linewidth}
    \begin{align*}
      \frac{\mstate \stepto \mstate'}{\mstate \stepto \Gamma(\mstate')}
    \end{align*}
  \end{minipage}
  \begin{minipage}{0.65\linewidth}
    \begin{align*}
      \Gamma(\tpl{\mexpr,\menv,\mstore,\mkont, \mktab, \mmemo}) &= \tpl{\mexpr,\menv,\mstore|_{\reaches(\mathit{root}, \mstore)}, \mkont, \mktab, \mmemo} \\
      \text{where } \mathit{root} &= \touchesk{\mkont}(\mktab) \cup
      \mathit{range}(\menv|_{\mathit{fv}(\mexpr)})
    \end{align*}
  \end{minipage}
\end{center}
%
The ``reaches'' metafunction ($\reaches$) is a transitive closure of ``touches'' as it weaves through the store, starting from some root set of addresses:

@@

\begin{enumerate}
\item{stack allocation for some bindings in an additional $\mframe \in \Store$}
\item{strong updates on stack frames for resolved addresses}
\end{enumerate}
There are two orthogonal features of the semantics, embodied by our semantics' two metafunctions ($\bind$ and $\lookup$, respectively in \autoref{fig:frame-semantics}).

@@

%%
The stack frame introduces a refinement over the original CESK machine, so we first prove it correct with respect to the concrete semantics, in a similar fashion to PDCFA.
%
\begin{theorem}[Correctness of stack frame refinement]\label{thm:refinement}
  The CES$\xi$K machine has a stuttering bisimulation with the CESK machine with fresh allocation.
\end{theorem}
This follows from the invariant that if a stack frame maps $\maddr$ to $\mval$, then the store maps $\maddr$ to $\mval$, meaning that the special lookup function is equivalent to lookup directly in the store (exploits immutability).
%
We stutter past the stack frame replacement steps.
%
%That is, if we always allocated bindings in the heap.
%
%The additional escape information that allows us to elide allocation in the heap introduces a second level of reasoning, where the burden of proof falls on the escape analysis.
%
%The escape analysis adds information to use in the proof, that if $\mvar$ never escapes, then every reference to allocations of $\mvar$'s binding will always be able to use $\mframe$ for the lookup.
%
%With this proposition, we can prove that the CES$\xi$K machine that always allocates in the store has a bisimulation with the refined machine that exploits escape analysis to avoid unnecessary changes to the store.
%

>>>>

%%
Testing new ideas in analysis for improving precision or performance can be a difficult venture.
%
Premature abstraction can mask correctness issues with complicated machinery we employ in our analyses\footnote{Oh so many personal experiences}.
%
We contend that analysis machinery should have an executable concrete counterpart that maintains the meaning of the language for effective testing.
%
Abstraction should then be a simple process that is ``obviously correct.''
%
The framework we present in this paper is applicable in the concrete such that a point-wise abstraction leads to the pushdown analyses in the literature.
%%

@@

Constructing models and analyzing models are different processes.
%
The goal of this work is to simplify the \textit{program} $\to$ \textit{model} process.
%
First-order languages have the luxury of a fully-precise control flow graph that can be extracted from a program AST in linear time.
%
In a sense, the first-order world gets their models ``for free.''
%
Higher-order recursion schemes (HORS) are a powerful model for simply typed functional programs with good model-checking properties~\citep{dvanhorn:Ong2006ModelChecking}, but the \textit{program} $\to$ \textit{HORS} process is still narrowly applicable.
%
It is still a matter of mystery for HORS to support programming language features like mutable state and untyped computation.
%%
As a middle ground, we show a simple way to transform one's programming language semantics into what can be viewed as a pushdown model constructor.
%%

@@

CFA2 is an online version of a technique called ``summarization'' from Sharir and Pnueli \citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with PDCFA's crucial ``$\epsilon$-closure graph''.
%
The online quality of the algorithm is what makes it suitable for higher-order languages, since CFA essentially interprets the program for each discovered state to build the model on-the-fly.
%
Summarization algorithms need not be restricted to languages with well-bracketed calls and returns.
%
The technique is adaptable for higher precision in the common case but to still handle difficult cases such as first-class control.
%
This was shown for the \rackett{call/cc} operator by~\citet{ianjohnson:Vardoulakis2011Pushdown}.
%
This work illuminated the fact that we can harness the enhanced technology of pushdown analyses in non-pushdown models of computation.
%
Doing this sacrifices call/return matching in the general case, but in practice the precision is much better than the alternative regular model that, \eg, 0-CFA would provide.
%%

@@

%%
The second piece of machinery serves two purposes:
%
(1) the state space contains all function callsites to project out a control-flow graph, and %
(2) the worklist semantics of execution makes it possible to appeal to an incomplete memo table, so the saved stacks allow new results to flow to calls that used incomplete memo information.
%
Once the machine semantics is in this form, simple point-wise abstraction leads to the summarization algorithms that we see in the literature.

\paragraph{The problem with incomplete memo information}
Consider a non-deterministic semantics where a function call reaches a result and memoizes, a second call uses the memo table (storing its stack), and then the original call reaches another result (since the worklist allows for such parallel reduction).
%
The second result had better also flow to the second call to maintain soundness!
%%

%%
\paragraph{Overview}
In \autoref{sec:pdcfa} we derive a cousin of PDCFA.
%
The following section shows how to extend PDCFA with abstract garbage collection. % \autoref{sec:gc}: we show how abstract garbage collection ``falls out'' of this approach.
%
Then in \autoref{sec:cfa2} we show the fundamental difference between PDCFA and CFA2 without first-class control. % : we make additions to the PDCFA semantics to get a direct-style CFA2 without first-class control}
%
\ifwcm{We show in \autoref{sec:wcm} that semantics with stack-inspection (beyond GC) also fit within this interpretation.}
%
Finally, we show in \autoref{sec:sr} that even though first-class control operators don't fit in the pushdown model, we can gracefully loosen the abstraction to analyze delimited, composable first-class control.

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}%{r|ll}
    \hline
% variable lookup
    $\tpl{\svar\mvar, \menv, \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
%    & \textsc{[variable lookup]}
    \\
% application
    $\tpl{\sapp{\mexpri0}{\mexpri1}, \menv, \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mexpri0, \menv, \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}, \mktab, \mmemo}$
%    & \textsc{[function eval]}
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{\mexpr, \menv, \mstore, \kcons{\kfn{\mval}}{\mkont}, \mktab, \mmemo}$
%    & \textsc{[argument eval]}
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\slam{\mvar}{\mexpr},\menv}}{\mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \mktab',
          \mmemo}$
%    & \textsc{[un-memoized call]}
     \\ & \quad if $(\mpoint, \mstore') \notin \dom(\mmemo)$
\\
    & or \\
    & $\tpl{\mval_\mathit{result},
            \mstore'',
            \mkont,
            \mktab',
            \mmemo}$ % & \textsc{[memo lookup]}
    \\ & \quad if $(\mval_\mathit{result},\mstore'') \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
          & $\mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{\mkont}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mctx}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, M'}$
%    & \textsc{[memoize/return]}
    \\ & \quad if $\mkont \in \mktab(\mctx)$
    \\ where & $\mmemo' = \joinone{\mmemo}{\mctx}{(\mval,\mstore)}$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

This semantics is only correct for fresh allocation strategies, which in turn make the reduction relation deterministic.
%
As it is, the CESK machine approximates the tabular machine, but not the other way around;
%
if we use a finite allocation strategy, then we don't explore some non-deterministic paths that lead to different results for a function call.
%
We thus explore all paths and share $\mktab$ and $\mmemo$ between all paths.

@@

{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mastate \in \sa{State} &::= \tpl{\mexpr,\menv,\mstore,\mkont} \\
  \System &= \wp(\sa{State}) \times \wp(\sa{State} \times \sa{State}) \times \KTab \times \Memo \\
  {\mathcal F} &: \sa{\State} \to \System \to \System
\end{align*}
\begin{align*}
  {\mathcal F}(\mastate_0)(\emptyset, R, \mktab, \mmemo) &= (\emptyset, R, \mktab, \mmemo) \\
  {\mathcal F}(\mastate_0)(F\uplus\set{\mastate}, R, \mktab, \mmemo) &= (F\cup F', R \cup R', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\wn(\mastate,\mktab,\mmemo) \stepto \mstate'} \\
                \wn(\tpl{\mexpr,\menv,\mstore,\mkont},\mktab,\mmemo)
                   &= \tpl{\mexpr,\menv,\mstore,\mkont,\mktab,\mmemo}
\end{align*}
\begin{align*}
    R' &= \setbuild{(\mastate,\mastate')}{\wn(\mastate', \_, \_) \in I \wedge \mastate\nequiv\tpl{\mval,\mstore,\krt{\mctx}}} \\
    \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \mktab', \_) \in I} \\
    \mmemo' &= \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \mmemo') \in I} \\
    F' &= (\setbuild{\mastate'}{\wn(\mastate', \_, \_) \in I} \cup \set{\mastate_0})\setminus (\pi_1 R \cup \pi_2 R)
  \end{align*}
}
%%

@@

Garbage collection ($\Gamma$) restricts the heap to just the reachable addresses, so a semantics with GC introduces the following reduction rule:
\begin{center}
  \begin{minipage}{0.3\linewidth}
    \begin{align*}
      \mstate \stepto \Gamma(\mstate)
    \end{align*}
  \end{minipage}
  \begin{minipage}{0.65\linewidth}
    \begin{align*}
      \Gamma(\tpl{\mexpr,\menv,\mstore,\mkont, \mktab, \mmemo}) &= \tpl{\mexpr,\menv,\mstore|_{\reaches(\mathit{root}, \mstore)}, \mkont, \mktab, \mmemo} \\
      \text{where } \mathit{root} &= \touchesk{\mkont}(\mktab) \cup
      \mathit{range}(\menv|_{\mathit{fv}(\mexpr)})
    \end{align*}
  \end{minipage}
\end{center}
%
The ``reaches'' metafunction ($\reaches$) is a transitive closure of ``touches'' as it weaves through the store, starting from some root set of addresses:

@@

\begin{enumerate}
\item{stack allocation for some bindings in an additional $\mframe \in \Store$}
\item{strong updates on stack frames for looked-up addresses}
\end{enumerate}
There are two orthogonal features of the semantics, embodied by our semantics' two metafunctions ($\bind$ and $\lookup$, respectively in \autoref{fig:frame-semantics}).

@@

%%
The stack frame introduces a refinement over the original CESK machine, so we first prove it correct with respect to the concrete semantics, in a similar fashion to PDCFA.
%
\begin{theorem}[Correctness of stack frame refinement]\label{thm:refinement}
  The CES$\xi$K machine has a stuttering bisimulation refinement with the CESK machine with fresh allocation.
\end{theorem}
This follows from the invariant that if a stack frame maps $\maddr$ to $\mval$, then the store maps $\maddr$ to $\mval$, meaning that the special lookup function is equivalent to lookup directly in the store (exploits immutability).
%
We stutter past the stack frame replacement steps.
%
%That is, if we always allocated bindings in the heap.
%
%The additional escape information that allows us to elide allocation in the heap introduces a second level of reasoning, where the burden of proof falls on the escape analysis.
%
%The escape analysis adds information to use in the proof, that if $\mvar$ never escapes, then every reference to allocations of $\mvar$'s binding will always be able to use $\mframe$ for the lookup.
%
%With this proposition, we can prove that the CES$\xi$K machine that always allocates in the store has a bisimulation with the refined machine that exploits escape analysis to avoid unnecessary changes to the store.
%
